\LARGE\textbf{Designing Studies and Evaluating Research Results:\\ Type M and Type S Errors for\\ Pearson Correlation Coefficient }

### 2021-02-19 Documents/Paper_main/Paper_main.Rnw
review editor

Why do these errors matter? The concepts of Type M and Type S errors allow enhancing researchers’ awareness of a complex process such as statistical inference. It gives a further tool that can be used to interpret and understand the important role of statistical power both when designing a new study or when evaluating previous results from the literature. In particular, it highlights the unwanted (and often overlooked) consequences on effect estimation when filtering for statistical significance in underpowered studies. The exaggeration of effect sizes, in the right or the wrong direction, has important implications on a theoretical and applied level. On a theoretical level, studies’ designs with high Type M and Type S errors can foster distorted expectations on the effect under study, triggering a vicious cycle for the planning of future studies. This point is relevant also for the design of replication studies, where the possible inflation of the original effect should be taken into account together with the smallest effect size of interest to limit the chances of being underpowered \parencite{buttonPowerFailureWhy2013}. When studies are used to inform policymaking and real-world interventions, implications can go beyond the academic research community and can impact society at large. In these settings, we could assist to a “hype and disappointment cycle” \parencite{gelmanOverconfidenceResearchCertainty2019}, where true effects turn out to be much less impressive than expected. This can produce undesirable consequences on people’s lives, a consideration that invites researchers to assume responsibility in effectively communicating the risks related to effects quantification.

@@

On a concluding note, we would like to clarify the relationship between publication bias and Type M and Type S errors. While the two are related, they operate at two different levels. Publication bias refers to a publication system that favours statistically significant results over non-statistically significant findings. This phenomenon alone cannot explain the presence of exaggerated effects. Imagine if all studies in the literature were conducted with high statistical power, then statistically significant findings would probably not be so extreme. The problem of exaggerated effect sizes in the literature can be explained only by a combination of publication bias with studies that have low statistical power. As previously shown, statistical power and Type M and Type S errors are related to each other: low statistical power corresponds to higher Type M and Type S errors. In the same way as statistical power is a characteristic of a study design, so are Type M and Type S, however, the two are qualitatively different in terms of the kind of reasoning they favour.

Statistical power is defined in terms of probability of rejecting the Null hypothesis and, even though this is based on an effect size of interest, the relationship “low power - high possibility of exaggeration” may not be straightforward for everyone. Instead, Type M and Type S errors directly quantify the possible exaggeration. Furthermore, their consideration protects against another possible pitfall. It is quite widespread the knowledge that even the smallest effects can become statistically significant if the sample size is large enough. However, if a statistically significant result is found paired with a large effect size, the results may be considered robust. However, this is not true in all cases. Here, the missing piece of information is  statistical power. If power is considered, researchers would realize that a large effect was found in a context where there was a low probability to find it. What is the meaning? This interpretation is not explicitly stating an important consequence: the study may find effects that are overestimated, if statistically significant. On the contrary, this becomes immediately clear once Type M and Type S errors are considered retrospectively. Similarly, considering Type M and Type S prospectively favours reason in terms of effect size rather than the probability of rejecting the null hypothesis when setting the sample size in a design analysis.

>>>>

Why do these errors matter? The concepts of Type M and Type S errors allow enhancing researchers’ awareness of a complex process such as statistical inference. Strictly speaking, Design Analysis does not add anything to the classical power analysis in the design of a study, since to a given level of power there is a corresponding Type M error. However, it is a valuable conceptual framework that can help researchers to understand the important role of statistical power both when designing a new study or when evaluating previous results from the literature. In particular, it highlights the unwanted (and often overlooked) consequences on effect estimation when filtering for statistical significance in underpowered studies. In these scenarios, there is not only a lower probability of finding a significant result but, even more importantly, any significant result would most likely lead to a misleading overestimation of the actual effect. The exaggeration of effect sizes, in the right or the wrong direction, has important implications on a theoretical and applied level. On a theoretical level, studies’ designs with high Type M and Type S errors can foster distorted expectations on the effect under study, triggering a vicious cycle for the planning of future studies. This point is relevant also for the design of replication studies, where the possible inflation of the original effect should be taken into account together with the smallest effect size of interest to limit the chances of being underpowered \parencite{buttonPowerFailureWhy2013}. When studies are used to inform policymaking and real-world interventions, implications can go beyond the academic research community and can impact society at large. In these settings, we could assist to a “hype and disappointment cycle” \parencite{gelmanOverconfidenceResearchCertainty2019}, where true effects turn out to be much less impressive than expected. This can produce undesirable consequences on people’s lives, a consideration that invites researchers to assume responsibility in effectively communicating the risks related to effects quantification.

@@

%----------------------------------------------------------------------------------%
%-----------------    Publication Bias and Significance Filter    -----------------%
%----------------------------------------------------------------------------------%

\subsection{Publication Bias and Significance Filter}

On a concluding note, we would like to clarify the relationship of Design Analysis with publication bias and statistical significance filter.

While publication bias and Type M and Type S errors are related, they operate at two different levels. Publication bias refers to a publication system that favours statistically significant results over non-statistically significant findings. This phenomenon alone cannot explain the presence of exaggerated effects. Imagine if all studies in the literature were conducted with high statistical power, then statistically significant findings would probably not be so extreme. The problem of exaggerated effect sizes in the literature can be explained only by a combination of publication bias with studies that have low statistical power. As previously shown, statistical power and Type M and Type S errors are related to each other: low statistical power corresponds to higher Type M and Type S errors.

The critical element is the application of the statistical significance filter without taking into account statistical power. Design Analysis per se does not solve this issue but, instead, it allows to recognize its problematic consequences. In the same way as statistical power is a characteristic of a study design, so are Type M and Type S, however, the two are qualitatively different in terms of the kind of reasoning they favour. Statistical power is defined in terms of probability of rejecting the Null hypothesis and, even though this is based on an effect size of interest, the relationship “low power - high possibility of exaggeration” may not be straightforward for everyone. Instead, Type M and Type S errors directly quantify the possible exaggeration. Furthermore, their consideration protects against another possible pitfall. It is quite widespread the knowledge that even the smallest effects can become statistically significant if the sample size is large enough. However, if a statistically significant result is found paired with a large effect size, the results may be considered robust. However, this is not true in all cases. Here, the missing piece of information is  statistical power. If power is considered, researchers would realize that a large effect was found in a context where there was a low probability to find it. What is the meaning? This interpretation is not explicitly stating an important consequence: the study may find effects that are overestimated, if statistically significant. On the contrary, this becomes immediately clear once Type M and Type S errors are considered retrospectively. Similarly, considering Type M and Type S prospectively favours reasoning in terms of effect size rather than the probability of rejecting the null hypothesis when setting the sample size in a design analysis.
